\documentclass[letterpaper]{IEEEtran}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage[capitalize]{cleveref}

% Convenience
\newcommand{\pfail}{P(\text{fail})}
\newcommand{\rho}{\varrho}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\etal}{\textit{et al.}}

% TODO marker
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

\begin{document}

\title{Cross-Corruption Safety Monitoring for Vision-Based\\Robot Manipulation via Adversarial Envelope Analysis}

\author{Julian~Quick%
\thanks{Manuscript received XX; revised XX. This work was conducted as an independent research project.}%
\thanks{J.~Quick is with \todo{affiliation}. E-mail: \todo{email}.}%
\thanks{Digital Object Identifier \todo{DOI}}%
}

\markboth{IEEE Robotics and Automation Letters, Vol.~XX, No.~XX, Month~2026}%
{Quick: Cross-Corruption Safety Monitoring for Vision-Based Robot Manipulation}

\maketitle

% ============================================================================
\begin{abstract}
Vision-language-action (VLA) policies for robot manipulation are brittle under camera occlusion---fingerprints, glare, and rain can degrade success rates from 100\% to near zero.
We present a lightweight safety monitor that predicts manipulation failure probability from a single camera frame, \emph{without requiring the occlusion type to have been seen during training}.
Our approach combines (i)~adversarial worst-case search via CMA-ES to characterize safe operating envelopes across occlusion types and severities, and (ii)~a 33K-parameter CNN head on a frozen ImageNet-pretrained ResNet-18 backbone trained on the resulting labeled episodes.
The key finding is that frozen ImageNet features encode corruption-agnostic visual degradation signals---contrast loss, blur, brightness shifts---that transfer across physically distinct occlusion mechanisms.
A systematic leave-one-out analysis across nine corruption types spanning six physical mechanisms --- lens contact (fingerprint, rain), optical (glare, defocus blur), blur (motion blur), sensor noise, atmospheric (fog), illumination (low-light), and digital (JPEG compression) --- demonstrates that the monitor achieves \todo{mean $\rho$ = X.XX $\pm$ X.XX} on held-out corruption types it has never encountered.
We additionally characterize how generalization improves with training corruption diversity and identify which corruption categories are easiest and hardest to predict out-of-distribution.
\end{abstract}

\begin{IEEEkeywords}
Safety monitoring, camera occlusion, vision-language-action models, SOTIF, adversarial testing, transfer learning, robot manipulation.
\end{IEEEkeywords}

% ============================================================================
\section{Introduction}
\label{sec:intro}

Vision-language-action (VLA) models such as InternVLA~\cite{internvla}, Octo~\cite{octo}, and RT-2~\cite{rt2} have demonstrated impressive manipulation capabilities, but their reliance on camera observations makes them vulnerable to perceptual degradation.
Recent benchmarking~\cite{liberoplus} reveals that common camera corruptions---fingerprints, glare, rain---can collapse VLA success rates from over 90\% to below 10\%, even at moderate severity levels.

This fragility is a core concern of the Safety of the Intended Functionality (SOTIF, ISO~21448~\cite{iso21448}) standard, which addresses failures arising not from hardware faults but from limitations of perception and decision-making under real-world conditions.
Camera occlusion is a canonical SOTIF triggering condition for vision-based systems: it is ubiquitous in deployment (warehouse dust, outdoor rain, operator fingerprints on lenses) yet difficult to anticipate exhaustively.

A practical safety monitor must therefore satisfy two requirements: (1)~it must detect when occlusion severity is likely to cause policy failure, and (2)~it must generalize to occlusion types \emph{not encountered during training}, since enumerating all possible degradation modes is infeasible.

We address both requirements with a two-stage approach.
First, we use adversarial optimization via CMA-ES~\cite{cmaes} to systematically find worst-case occlusion parameters at each severity (budget) level, producing a \emph{safe operating envelope} that maps occlusion coverage to certified success rates (\cref{sec:envelope}).
Second, we train a lightweight CNN safety monitor on the resulting labeled episodes and evaluate its ability to generalize to held-out corruption types (\cref{sec:safety_predictor}).

Our key finding is that a frozen ImageNet-pretrained ResNet-18 backbone, paired with a small trainable head (33K parameters), learns corruption-agnostic failure signals that transfer across physically distinct occlusion mechanisms.
Trained on fingerprint and glare corruptions, the monitor achieves strong rank correlation with failure probability under rain occlusion it has never seen.

\noindent\textbf{Contributions:}
\begin{enumerate}
    \item A systematic leave-one-out evaluation across nine corruption types spanning six physical mechanisms, demonstrating corruption-agnostic failure detection.
    \item A lightweight cross-corruption safety monitor (33K trainable parameters) that generalizes to unseen occlusion types via frozen pretrained features.
    \item A corruption diversity curve showing how many training corruption types suffice for effective OOD generalization.
    \item An adversarial envelope analysis framework using CMA-ES to characterize worst-case policy performance under parametric occlusion.
\end{enumerate}

% ============================================================================
\section{Related Work}
\label{sec:related}

\subsection{Robustness of Vision-Based Manipulation Policies}

Recent work has exposed the fragility of learned manipulation policies under visual corruptions.
LIBERO-Plus~\cite{liberoplus} systematically benchmarks VLA robustness, finding dramatic performance drops under ImageNet-C-style corruptions.
RobustVLA~\cite{robustvla} proposes robustness-aware reinforcement learning post-training to harden policies.
FPC-VLA~\cite{fpcvla} combines failure prediction with corrective actions.
CRT~\cite{crt} restores corrupted visual inputs before feeding them to VLA policies.
These approaches either make the policy more robust or restore corrupted inputs; our work is complementary---we build an external safety monitor that detects when the policy is likely to fail, regardless of whether the policy or input has been hardened.

\subsection{Runtime Safety Monitoring}

Sentinel~\cite{sentinel} monitors generative manipulation policies at runtime by detecting behavioral anomalies at the policy output level.
Ma~\etal~\cite{ma2024} predict failure from camera faults using multimodal features with 4000 real-world samples.
FORTRESS~\cite{fortress} addresses real-time out-of-distribution failure prevention.
Our approach differs in monitoring at the \emph{perceptual input level} and demonstrating zero-shot generalization to unseen corruption types---a property not evaluated in prior work.

\subsection{Adversarial Testing for Safety}

Adversarial testing has been used to find failure modes in autonomous driving~\cite{adv_driving} and manipulation~\cite{adv_manip}.
CMA-ES is well-suited for this setting as it is derivative-free, handles moderate-dimensional parameter spaces efficiently, and naturally parallelizes across population members.
We extend adversarial testing to produce certified safe operating envelopes parameterized by occlusion severity.

\subsection{SOTIF and Safety Standards}

ISO~21448~\cite{iso21448} defines a framework for analyzing safety of the intended functionality, distinguishing known/unknown safe and unsafe scenarios.
Our envelope analysis maps directly to SOTIF's scenario classification: adversarially tested conditions become ``known unsafe,'' while cross-corruption generalization moves previously ``unknown unsafe'' scenarios toward ``known'' status.
\todo{Expand SOTIF mapping in discussion.}

% ============================================================================
\section{Problem Formulation}
\label{sec:problem}

Consider a vision-based manipulation policy $\pi$ that maps camera observations $o_t \in \mathbb{R}^{H \times W \times 3}$ to actions $a_t$.
An occlusion model $\mathcal{O}_\theta: \mathbb{R}^{H \times W \times 3} \to \mathbb{R}^{H \times W \times 3}$ parameterized by $\theta \in \Theta$ degrades the observation at each timestep: $\tilde{o}_t = \mathcal{O}_\theta(o_t, t)$.
A budget function $b(\theta) \in [0, 1]$ measures the fraction of pixels significantly affected by the occlusion.

\noindent\textbf{Safe Operating Envelope.}
For a given occlusion type $\mathcal{O}$ and budget level $\beta$, we seek the worst-case success rate:
\begin{equation}
    \text{SR}^*(\mathcal{O}, \beta) = \min_{\theta : b(\theta) \leq \beta} \; \mathbb{E}\left[\text{success}(\pi, \mathcal{O}_\theta)\right]
\label{eq:envelope}
\end{equation}
The safe operating envelope is the set of $(\mathcal{O}, \beta)$ pairs where $\text{SR}^*(\mathcal{O}, \beta) \geq \tau_{\text{safe}}$.

\noindent\textbf{Safety Monitor.}
We train a monitor $f_\phi: \mathbb{R}^{H \times W \times 3} \to [0, 1]$ that maps a single (potentially occluded) frame to a predicted failure probability $\hat{p} = f_\phi(\tilde{o}_t)$.
The monitor is trained on corruption types $\mathcal{O}_{\text{train}}$ and evaluated on held-out types $\mathcal{O}_{\text{test}}$ to assess cross-corruption generalization.

% ============================================================================
\section{Parametric Occlusion Models}
\label{sec:occlusion}

We organize visual corruptions by physical mechanism following the taxonomy of Hendrycks \& Dietterich~\cite{hendrycks2019}, adapted for fixed-mount robot cameras.
Our set spans six categories to support systematic cross-corruption generalization analysis.
All corruptions are applied as image-space post-processing on rendered camera frames and are parameterized with a budget level $\beta \in [0, 1]$ controlling severity.

\cref{tab:corruptions} summarizes the nine corruption types used in our experiments.

\begin{table}[t]
\centering
\caption{Corruption taxonomy organized by physical mechanism.}
\label{tab:corruptions}
\footnotesize
\begin{tabular}{llrl}
\toprule
\textbf{Category} & \textbf{Corruption} & \textbf{Params} & \textbf{Source} \\
\midrule
Lens contact  & Fingerprint smudge & 24 & Gu~\etal~\cite{gu2009} \\
Lens contact  & Rain drops         &  5 & \cite{camera_occlusion} \\
Optical       & Glare / lens flare &  6 & Physically-inspired \\
Optical       & Defocus blur       &  1 & ImageNet-C~\cite{hendrycks2019} \\
Blur          & Motion blur        &  2 & ImageNet-C~\cite{hendrycks2019} \\
Sensor        & Gaussian noise     &  2 & ImageNet-C~\cite{hendrycks2019} \\
Atmospheric   & Fog / haze         &  2 & ImageNet-C~\cite{hendrycks2019} \\
Illumination  & Low-light          &  2 & LIBERO-Plus~\cite{liberoplus} \\
Digital       & JPEG compression   &  1 & ImageNet-C~\cite{hendrycks2019} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Physics-Based Models}

\noindent\textbf{Fingerprint smudge.}
Based on the thin-film optics model of Gu~\etal~\cite{gu2009}, we simulate oil-film residue from finger contact.
Each fingerprint is parameterized by center, angle, scale, ridge frequency, amplitude, opacity, and smear angle (8 parameters per print).
The rendering pipeline generates a whorl-pattern ridge mask, applies defocus blur, and composites via attenuation and veiling glare.

\noindent\textbf{Glare / lens flare.}
We model lens flare as diffraction spikes, haze, chromatic aberration, and lens ghosts.
Compositing uses screen blending: $I_{\text{out}} = 1 - (1 - I_{\text{base}})(1 - I_{\text{flare}})$.

\noindent\textbf{Rain drops.}
Refraction-based raindrop model~\cite{camera_occlusion} parameterized by drop density, radius range, magnification, and blur.

\subsection{Standard Corruption Models}

Following ImageNet-C~\cite{hendrycks2019} and LIBERO-Plus~\cite{liberoplus}, we include:
\textbf{Gaussian noise} (additive sensor read noise),
\textbf{JPEG compression} (block artifacts at controllable quality),
\textbf{motion blur} (directional kernel with variable angle and size),
\textbf{defocus blur} (Gaussian blur with variable sigma),
\textbf{fog} (uniform alpha blend with white), and
\textbf{low-light} (gamma darkening with amplified sensor read noise, simulating reduced photon count).

\subsection{Budget Constraint}

All models enforce a budget level $\beta$ controlling corruption severity.
For spatially-localized models (fingerprint, rain), budget is the fraction of pixels above a dirty threshold, enforced via binary search.
For global models (fog, low-light, blur), budget linearly scales the corruption intensity parameter.
This enables fair comparison across corruption types at matched severity levels.

\todo{Add figure showing all nine corruption types at matched budget levels.}

% ============================================================================
\section{Adversarial Envelope Analysis}
\label{sec:envelope}

\subsection{CMA-ES Optimization}

We use the Covariance Matrix Adaptation Evolution Strategy (CMA-ES)~\cite{cmaes} to solve \cref{eq:envelope} for each occlusion type and budget level.
CMA-ES is well-suited because: (i)~occlusion application is non-differentiable, (ii)~the parameter space is moderate-dimensional (6--24 dimensions), and (iii)~population-based evaluation naturally parallelizes across simulation instances.

At each generation, CMA-ES proposes a population of candidate occlusion parameters $\{\theta_1, \ldots, \theta_\lambda\}$.
Each candidate is evaluated by running $N$ episodes with the occluded policy and computing the success rate.
The fitness (negative success rate) is reported back to CMA-ES, which updates its search distribution.
After convergence, the best candidate undergoes a final evaluation with $N_{\text{final}} = 25$ episodes.

\subsection{Budget Projection}

To ensure fair comparison across budget levels, all occlusion models implement a budget projection step.
Given a target budget $\beta$, we perform binary search over a global scale factor that modulates occlusion intensity until the pixel coverage constraint $b(\theta) \leq \beta$ is satisfied.
This decouples the spatial structure of the occlusion (optimized by CMA-ES) from its intensity (set by the budget).

\subsection{Envelope Classification}

We classify each $(\mathcal{O}, \beta)$ condition into zones:
\begin{itemize}
    \item \textbf{Safe}: $\text{SR}^*(\mathcal{O}, \beta) \geq 0.95$
    \item \textbf{Marginal}: $0.70 \leq \text{SR}^*(\mathcal{O}, \beta) < 0.95$
    \item \textbf{Unsafe}: $\text{SR}^*(\mathcal{O}, \beta) < 0.70$
\end{itemize}
The \emph{maximum safe budget} is the largest $\beta$ for which the condition remains safe.
The \emph{breakdown budget} is the smallest $\beta$ at which the condition becomes unsafe.

\todo{Add envelope results table and figure.}

% ============================================================================
\section{Cross-Corruption Safety Monitor}
\label{sec:safety_predictor}

\subsection{Architecture}

Our safety monitor uses a ResNet-18~\cite{resnet} backbone pretrained on ImageNet~\cite{imagenet}, with all backbone parameters frozen.
A lightweight classification head maps the 512-dimensional feature vector to a scalar failure logit:
\begin{equation}
    f_\phi(o) = \sigma\!\left(W_2 \cdot \text{ReLU}(W_1 \cdot g(o) + b_1) + b_2\right)
\end{equation}
where $g(\cdot)$ denotes the frozen ResNet-18 feature extractor, $W_1 \in \mathbb{R}^{64 \times 512}$, $W_2 \in \mathbb{R}^{1 \times 64}$, and $\sigma$ is the sigmoid function.
Dropout ($p = 0.3$) is applied after the ReLU.
The head contains approximately 33K trainable parameters.

\subsection{Training}

\noindent\textbf{Data collection.}
For each training corruption type (fingerprint, glare) and budget level $\beta \in \{0.1, 0.3, 0.5, 0.7, 0.9\}$, we collect episodes with both random and adversarially optimized occlusion parameters.
Each episode yields sampled frames labeled with the episode outcome (success/failure).

\noindent\textbf{Loss function.}
We use binary cross-entropy with logits and class weighting to handle imbalanced data:
\begin{equation}
    \mathcal{L} = -\frac{1}{N}\sum_{i=1}^{N} \left[w_+ y_i \log \hat{p}_i + (1 - y_i) \log (1 - \hat{p}_i)\right]
\end{equation}
where $w_+ = n_{\text{success}} / n_{\text{failure}}$ upweights the minority failure class.

\noindent\textbf{Episode-level splitting.}
To prevent data leakage from temporally correlated frames within the same episode, we split at the episode level: all frames from a given episode appear exclusively in either the training or validation set ($80/20$ split).

\noindent\textbf{Optimization.}
Adam optimizer with learning rate $10^{-3}$, batch size 32, for 50 epochs with ReduceLROnPlateau scheduling.

\subsection{Leave-One-Out Cross-Corruption Protocol}
\label{sec:loo}

To systematically evaluate cross-corruption generalization, we employ a leave-one-out (LOO) protocol over all nine corruption types.
In each fold $k$, one corruption type $\mathcal{O}_k$ is held out.
The safety monitor is trained on episodes collected from the remaining eight types and evaluated on the held-out type.

For each held-out type and budget level, we run $N$ evaluation episodes and compute:
\begin{itemize}
    \item The Spearman rank correlation $\rho$ between budget level and mean predicted $\hat{p}_{\text{fail}}$ (severity ranking quality).
    \item The episode-level AUROC for binary failure prediction (discriminative power).
    \item The Expected Calibration Error (ECE) from reliability diagrams.
\end{itemize}
Bootstrapped 95\% confidence intervals (1000 resamples) are reported for all metrics.

\noindent\textbf{Corruption diversity curve.}
We additionally vary the training set size $|\mathcal{O}_{\text{train}}| \in \{2, 4, 6, 8\}$ by randomly sampling subsets and evaluating on the remaining types.
This reveals how many corruption types are needed for effective generalization.

% ============================================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}

\noindent\textbf{Simulation.}
We use SimplerEnv~\cite{simplerenv} built on SAPIEN/ManiSkill2 for policy evaluation.
Corruptions are applied as post-processing on rendered camera frames at each simulation step.
Object positions are randomized at each episode reset.

\noindent\textbf{Policy.}
InternVLA-M1~\cite{internvla}, a vision-language-action model, serves as the manipulation policy.
On the \texttt{google\_robot\_pick\_coke\_can} task, it achieves 100\% success rate (50/50 episodes) under clean conditions.

\noindent\textbf{Data collection.}
For each of the nine corruption types (\cref{tab:corruptions}) at budget levels $\beta \in \{0.1, 0.3, 0.5, 0.7, 0.9\}$, we collect 10 episodes with random occlusion parameters, yielding ${\sim}450$ episodes total.
Each episode contributes sampled frames labeled with the episode outcome (success/failure).
This shared dataset is partitioned differently for each LOO fold, making the CNN training cost negligible (33K parameters, ${\sim}$2 minutes per fold).

\subsection{Motivating Experiment}

In preliminary experiments, we trained the safety monitor on only fingerprint and glare corruptions and evaluated on held-out rain.
A from-scratch CNN achieved $\rho = -0.62$ (anti-correlated), while the frozen ResNet-18 backbone achieved $\rho = 0.90$ ($p = 0.037$).
This surprising result --- a 152-percentage-point improvement from pretrained features alone --- motivated the systematic LOO analysis that follows.

\subsection{Leave-One-Out Results}

\begin{table}[t]
\centering
\caption{Leave-one-out cross-corruption generalization. Each row holds out one corruption type; the monitor is trained on the other eight. Higher $\rho$ and AUROC indicate better OOD generalization.}
\label{tab:loo}
\begin{tabular}{llccc}
\toprule
\textbf{Category} & \textbf{Held-out} & \textbf{$\rho$} & \textbf{95\% CI} & \textbf{AUROC} \\
\midrule
Lens contact  & Fingerprint & \todo{} & \todo{} & \todo{} \\
Lens contact  & Rain        & \todo{} & \todo{} & \todo{} \\
Optical       & Glare       & \todo{} & \todo{} & \todo{} \\
Optical       & Defocus     & \todo{} & \todo{} & \todo{} \\
Blur          & Motion blur & \todo{} & \todo{} & \todo{} \\
Sensor        & Gauss.\ noise & \todo{} & \todo{} & \todo{} \\
Atmospheric   & Fog         & \todo{} & \todo{} & \todo{} \\
Illumination  & Low-light   & \todo{} & \todo{} & \todo{} \\
Digital       & JPEG        & \todo{} & \todo{} & \todo{} \\
\midrule
\multicolumn{2}{l}{\textbf{Mean $\pm$ std}} & \todo{} & --- & \todo{} \\
\bottomrule
\end{tabular}
\end{table}

\todo{Fill in LOO results table from experiments. Discuss which corruption types are hardest/easiest to generalize to and why.}

\subsection{Corruption Diversity Curve}

\todo{Plot mean held-out $\rho$ vs.\ number of training corruption types $\{2, 4, 6, 8\}$ with error bars from random subsets. Discuss saturation point --- how many types are needed for effective generalization?}

\subsection{Safe Operating Envelope Results}

\todo{Insert envelope results table. Include adversarial vs.\ random SR, zone classifications, max safe budget, breakdown budget.}

\subsection{Ablation Studies}

\todo{%
Ablations:
\begin{itemize}
    \item Frozen vs.\ fine-tuned vs.\ from-scratch backbone
    \item ResNet-18 vs.\ ResNet-50 vs.\ ViT-B/16 backbone
    \item Training data size (learning curve)
\end{itemize}
}

\subsection{Baseline Comparisons}

\todo{%
Compare against BRISQUE, NIQE, pixel coverage, and autoencoder baselines using same LOO evaluation protocol.
}

% ============================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Why Cross-Corruption Transfer Works}

We hypothesize that the frozen ImageNet backbone encodes generic visual quality features---edge sharpness, contrast, texture coherence---that degrade similarly under physically distinct occlusion mechanisms.
Fingerprints attenuate contrast and add low-frequency smearing; glare saturates highlights and reduces dynamic range; rain introduces local blur and magnification.
All three reduce the information content available to the downstream policy, and the pretrained features capture this reduction without being specific to any corruption type.

\todo{Support with GradCAM visualization showing what the CNN attends to under each corruption type.}

\subsection{Practical Deployment Considerations}

The safety monitor runs inference on a single frame in under \todo{X}ms on a consumer GPU, making it suitable for real-time deployment alongside the manipulation policy.
The envelope analysis provides deployment-time guidance: operators can specify acceptable risk thresholds and the system reports the maximum occlusion coverage that remains within the certified safe zone.

\subsection{Limitations}

\begin{itemize}
    \item \textbf{Simulation only.} All experiments are conducted in simulation. Real-world camera occlusions may differ in appearance and statistics. \todo{Add real-world pilot results.}
    \item \textbf{Task scope.} We evaluate on pick-and-place tasks; generalization to contact-rich or long-horizon tasks is untested. \todo{Add additional tasks.}
    \item \textbf{Occlusion diversity.} Three corruption types may not span the full space of real-world degradation modes.
    \item \textbf{Calibration.} We evaluate ranking (Spearman $\rho$) but not probability calibration. \todo{Add ECE analysis.}
\end{itemize}

\subsection{Connection to SOTIF}

\todo{Map our framework formally to ISO 21448 scenario classification: known safe, known unsafe, unknown safe, unknown unsafe. Show how adversarial testing converts ``unknown unsafe'' to ``known unsafe'' and how cross-corruption generalization addresses ``unknown'' scenarios.}

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented a framework for safety monitoring of vision-based manipulation under camera occlusion, combining adversarial envelope analysis with a lightweight cross-corruption safety predictor.
The central finding is that frozen ImageNet features enable a 33K-parameter CNN head to predict failure under occlusion types never encountered during training, suggesting that practical safety monitors need not enumerate every possible degradation mode.
Adversarial CMA-ES search provides certified worst-case bounds, while the cross-corruption monitor provides real-time per-frame risk assessment.
\todo{Update conclusion with final results.}

% ============================================================================
\section*{Acknowledgments}
\todo{Acknowledgments.}

% ============================================================================
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
